Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-25 arxiv/proof_gaussian_case.tex
wip

%\input{proof_two_tasks}

@@

\medskip
\noindent{\bf Deriving the bias asymptotics.} Finally, we describe how to derive the bias asymptotics. We can write the left-hand side of equation \eqref{lem_cov_derv_eq} using the derivative of $\cal G$ with respect to $z$ at $z=0$. More precisely, using equation \eqref{eigen2extra} we can obtain that
\begin{align}
&n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}X^{(1)} + (X^{(2)})^{\top}X^{(2)} }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
&=n^2 w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\left(   \Lambda U^\top (Z^{(1)})^\top Z^{(1)} U\Lambda  + V^\top (Z^{(2)})^\top Z^{(2)}V\right)^{-2}V^\top\Sigma_2^{-1/2}\Sigma_1^{1/2}w \nonumber\\
&=  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\cal G'(0)V^\top\Sigma_2^{-1/2}\Sigma_1^{1/2}w,\label{calculate G'}
\end{align}
where we used equation \eqref{rewrite X as R} in the second step. Since the matrix limit of $\cal G(z)$ is given by equation \eqref{matrix limit}, it is natural to guess that the matrix limit of $\cal G'(0)$ is given by
\be\label{cal G'0}\cal G'(0) \approx \left.\frac{\dd}{\dd z}\right|_{z=0}(-z\id_p+a_{1}(z)\Lambda^2 + a_{2}(z)\id_p)^{-1} = \frac{\id_p- a_1'(0)\Lambda^2 - a_2'(0)\id_p}{(a_{1}(0)\Lambda^2 + a_{2}(0)\id_p)^2}.\ee
If we let $a_3:=-a_1'(0)$ and $a_4:=-a_2'(0)$, then taking implicit differentiation of equation \eqref{selfomega_a} we can check that $(a_3,a_4)$ satisfies equation \eqref{eq_a34extra}. Then inserting \eqref{cal G'0} into \eqref{calculate G'}, we obtain that
\begin{align}
& n^2\bignorm{\Sigma_2^{1/2} \bigbrace{ (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}) }^{-1} \Sigma_1^{1/2} w}^2 \nonumber\\
&\approx  w^\top \Sigma_1^{1/2}\Sigma_2^{-1/2}V\frac{a_3\Lambda^2 +(1+ a_4)\id_p}{(a_{1}\Lambda^2 + a_{2}\id_p)^2}V^\top \Sigma_2^{-1/2}\Sigma_1^{1/2}w= w^{\top} \Pi_\bias w, \label{calculatePibias}
\end{align}
where in the last step we used $M = \Sigma_1^{1/2}\Sigma_2^{-1/2}$ and $V \Lambda^2 V^\top=M^\top M$. This concludes equation \eqref{lem_cov_derv_eq}.
Note that in order to have the approximate identity for $\cal G'(0)$ in equation \eqref{cal G'0}, we not only need to know the asymptotics of $\cal G(0)$, but also need to know the asymptotics of $\cal G(z)$ for general $z$ around $z=0$. This is the main reason why we need to take a general $z$ in the definition of resolvents. The rigorous proof of equation \eqref{lem_cov_derv_eq} will be given in Section \ref{sec pf RMTlemma}, where we justify the approximate identity in equation \eqref{cal G'0}.
%In the above definition, we have taken the argument of $\cal G$ to be a general complex number, because we will need to use $\cal G'(0)$ in the proof of Lemma \ref{lem_cov_derivative}, which requires a good estimate of $\cal G(z)$ for $z$ around the origin.
%For the variance asymptotic limit, we study the resolvent
%	\[ R(z):= \bigbrace{\Sigma_2^{-1/2}( (X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)}))\Sigma_2^{-1/2} - z \id}^{-1}, \text{ for any } z\in \C \text{ around } z=0. \]
%Using the techniques from \citet{Anisotropic} and \citet{yang2019spiked}, we find the asymptotic limit of $R(z)$ for any $z$ as $p$ goes to infinity, denoted by $R_\infty(z)$, with an almost optimal convergence rate of $p$.
%In particular, when $z=0$, the asymptotic limit of equation \eqref{lem_cov_shift_eq} is given by
%	\[ \tr[\Sigma_2^{-1/2} \Sigma \Sigma_2^{-1/2}R_\infty(0)]. \]
%
%For the bias asymptotic limit, we show in \todo{where?} that
%$$\bignorm{\Sigma_2^{1/2} ((X^{(1)})^{\top}(X^{(1)}) + (X^{(2)})^{\top}(X^{(2)})^{-1} w}^2= w^\top \Sigma_2^{-1/2}R'(0)\Sigma_2^{-1/2} w.$$
%Hence its limit can be calculated through $R_\infty'(z)$, which gives the expression in \eqref{lem_cov_derv_eq}.
%We leave the full proof of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative} to Appendix \ref{sec_maintools}.
%Combining the above two results, we provide the proof of Theorem \ref{thm_main_informal} in Section \ref{app_proof_main_thm}.

The above arguments are the core of the main proof. To have a rigorous proof, we need to estimate each error carefully, and extend the Gaussian case to the more general case where the entries of $Z^{(1)}$ and $Z^{(2)}$ only satisfy certain moment assumptions. These will make the real argument rather tedious, but the methods we used are standard in the random matrix literature \cite{erdos2017dynamical,Anisotropic}. For the full rigorous proof, we refer the reader to Section \ref{appendix RMT}.

>>>>

%\input{proof_two_tasks}
The main goal of this section is to prove Theorem \ref{thm_main_RMT}.
%Recall that $n:=n_1+n_2$, which is the total number of data in the two tasks.
%Before giving the main proof, we first introduce some notations and tools.
%Following the notations in \cite{EKYY,EKYY1}, we will use the following definition to characterize events of overwhelming probability.
%
%\begin{definition}[overwhelming probability event] \label{high_prob}
%Define
%\begin{equation}\label{def_phi}
%\varphi:=(\log N)^{\log \log N}.
%\end{equation}
%We say that an $N$-dependent event $\Omega$ holds with {\it{$\xi$-overwhelming probability}} if there exist constant $c,C>0$ independent of $N$, such that
%\begin{equation}
%\mathbb{P}(\Omega) \geq 1-N^{C} \exp(- c\varphi^{\xi}),  \label{D25}
%\end{equation}
%for all sufficiently large $N$. For simplicity, for the case $\xi=1$, we just say {\it{overwhelming probability}}. Note that if (\ref{D25}) holds, then $\mathbb P(\Omega) \ge 1 - \exp(-c'\varphi^{\xi})$ for any constant $0\le c' <c$.
%\end{definition}
%\begin{remark}
%For any $c, C>0$, there exists a $0< c^{\prime}<c$, such that $N^{C} \exp(-c \varphi^{\xi}) \leq \exp(-c^{\prime} \varphi^{\xi})$. Hence, if
%\begin{equation}
%\mathbb{P}(\Omega) \geq 1- \exp(-c \varphi^{\xi})
%\end{equation}
%$\Omega$ is also a $\xi$-overwhelming probability event.
%\end{remark}
For simplicity of notations, we will use the following notion of stochastic domination, which was first introduced in \cite{Average_fluc} and subsequently used in many works on random matrix theory. It greatly simplifies the presentation of the results and their proofs by systematizing statements of the form ``$\xi$ is bounded by $\zeta$ with overwhelming probability up to a small power of $n$".

\begin{definition}[Stochastic domination]\label{stoch_domination}
%\begin{itemize}
%\item[(i)]
Let $\xi\equiv \xi^{(n)}$ and $\zeta\equiv \zeta^{(n)}$ be two $n$-dependent random variables.
%\begin{itemize} \item[(i)]
We say $\xi$ is stochastically dominated by $\zeta$, denoted by $\xi\prec \zeta$ or $\xi=\OO_\prec(\zeta)$, if for any (small) constant $\e>0$ and (large) constant $D>0$,
\[ \bbP\left(|\xi| >n^\e |\zeta|\right)\le n^{-D}\]
for large enough $n\ge n_0(\e, D)$. Moreover, if $\xi$ and $\zeta$ depend on a parameter $u$, then we say $\xi$ is stochastically dominated by $\zeta$ uniformly in $u\in \cal U$,  if for any constants $\e,D>0$,
\[\sup_{u\in \cal U}\bbP\left(|\xi(u)|>n^\e |\zeta(u)|\right)\le n^{-D}\]
for large enough $n\ge n_0(\e, D)$.
%\item[(ii)]
%\end{itemize}
%(i) Let
%\[\xi=\left(\xi^{(n)}(u):n\in\bbN, u\in U^{(n)}\right),\hskip 10pt \zeta=\left(\zeta^{(n)}(u):n\in\bbN, u\in U^{(n)}\right)\]
%be two families of nonnegative random variables, where $U^{(n)}$ is a possibly $n$-dependent parameter set. We say $\xi$ is stochastically dominated by $\zeta$, uniformly in $u$, if for any fixed (small) $\epsilon>0$ and (large) $D>0$,
%\[\sup_{u\in U^{(n)}}\bbP\left[\xi^{(n)}(u)>n^\epsilon\zeta^{(n)}(u)\right]\le n^{-D}\]
%for large enough $n\ge n_0(\epsilon, D)$, and we shall use the notation $\xi\prec\zeta$.
%%Throughout this paper, the stochastic domination will always be uniform in all parameters that are not explicitly fixed (such as matrix indices, and $z$ that takes values in some compact set).
%%Note that $N_0(\epsilon, D)$ may depend on quantities that are explicitly constant, such as $\tau$ in Assumption \ref{assm_big1} and equation \eqref{assm_gap}.
%If for some complex family $\xi$ we have $|\xi|\prec\zeta$, then we will also write $\xi \prec \zeta$ or $\xi=\OO_\prec(\zeta)$.
%\item[(ii)]
%(ii) We extend the definition of $\OO_\prec(\cdot)$ to matrices in the weak operator sense as follows. Let $A$ be a family of random matrices and $\zeta$ be a family of nonnegative random variables. Then $A=O_\prec(\zeta)$ means that $\left|\left\langle\mathbf v, A\mathbf w\right\rangle\right|\prec\zeta \| \mathbf v\|_2 \|\mathbf w\|_2 $ uniformly in any deterministic vectors $\mathbf v$ and $\mathbf w$. Here and throughout the following, whenever we say ``uniformly in any deterministic vectors", we mean that ``uniformly in any deterministic vectors belonging to a set of cardinality $N^{\OO(1)}$".
%\item[(iv)]
%\end{itemize}
\end{definition}
\begin{remark}
Note that since we allow an $n^\e$ factor in the definition of stochastic domination, we have the seemingly strange inequality $(\log n)^C\prec 1$ for any constant $C>0$.
\end{remark}

\begin{example}\label{example stoc}
Given a random variable $\xi$, if its moments exist up to any order, then we have $|\xi|\prec 1$. In fact, for any constants $\e, D>0$, by Markov's inequality we have
$$ \P(|\xi|\ge n^{\e})\le n^{-k\e}{\E |\xi|^k}\le n^{-D},$$
if we choose $k$ large enough such that $k\e>D$. As a special case, we have $|\xi|\prec 1$ for a Gaussian random variable $\xi$ with unit variance.
\end{example}

%\HZ{Why is this below part of the def. of stochastic domination?}
In the following proof, we say an event $\Xi$ holds with overwhelming probability if for any constant $D>0$, $\mathbb P(\Xi)\ge 1- n^{-D}$ for large enough $n$. Moreover, we say $\Xi$ holds with overwhelming probability in an event $\Omega$ if for any constant $D>0$, $\mathbb P(\Omega\setminus \Xi)\le n^{-D}$ for large enough $n$.

In this section, we shall prove a more fundamental random matrix result in a little more general setting than the one in Section \todo{add reference}; see Theorem \ref{LEM_SMALL} below. %\ref{sec_prelim}.
 Then Theorem \ref{thm_main_RMT} will follow from this result as immediate corollaries. We first state the exact assumptions needed for our purpose.

%We summarize our basic assumptions here for future reference. %Note that this assumption is in accordance with the assumptions of Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative}, except that we rescale the entries of $Z^{(1)}$ and $Z^{(2)}$ by $n^{-1/2}$. %and allow $\rho_1$ to be smaller than 1.
\begin{assumption}\label{assm_big1}
As in Section \todo{add reference}, we consider random matrices $X^{(1)}=Z^{(1)}\Sigma_1^{1/2}$ and $X^{(2)}=Z^{(2)}\Sigma_2^{1/2}$, where $Z^{(1)}$ and $Z^{(2)}$ are independent $n_1\times p$ and $n_2\times p$ random matrices with i.i.d. entries of zero mean and unit variance. Recall that $\rho_1= n_1/p$ and $\rho_2=n_2/p$. We assume that for a small constant $0<\tau<1$,
%\HZ{We have defined $\rho_1,\rho_2$ already in sec 2. Consider using ``Recall that $\rho_1 =, \rho_2 =$.''}
\be\label{assm2}
0\le \rho_1 \le \tau^{-1}, \quad 1+\tau \le \rho_{2} \le \tau^{-1}.
\ee
Recall that the singular value decomposition of $M = \Sig_1^{1/2} \Sig_2^{-1/2}$ is $U \Lambda V^{\top}$. We assume that
\begin{equation}\label{assm32}
\tau \le \lambda_p \le \lambda_1 \le \tau^{-1} .%, \quad \max\left\{\pi_A^{(n)}([0,\tau]), \pi_B^{(n)}([0,\tau])\right\} \le 1 - \tau .
\end{equation}
Finally, recalling the notations in equation \eqref{labelZ}, we assume that $Z^{(1)}$ and $Z^{(2)}$ satisfy
\begin{equation}
n^{-1/2}\max_{i\in \cal I_0,\mu \in \cal I_1}\vert Z^{(1)}_{\mu i}\vert \prec q,\quad n^{-1/2}\max_{i\in \cal I_0,\nu \in \cal I_2}\vert Z^{(2)}_{\nu i}\vert \prec q, \label{eq_support}
\end{equation}
for some deterministic parameter $q$ satisfying $ n^{-{1}/{2}} \leq q \leq n^{- \tau} $ for a small constant $\tau>0$.
%Here we recall the notations in \eqref{labelZ}.
 %and (\ref{assm2}). We assume that $T$ is an $M\times M$ deterministic diagonal matrix satisfying (\ref{simple_assumption}) and (\ref{assm3}).
\end{assumption}
%\HZ{This is restating what's stated above again. Remove redundant statements (unless there's something critical missing in Sec 2).} [\cor FY: The setting in sec 1 is too vague for the purpose of this section in the sense that it is hard to find the exact properties we need. In this section, I want to state rigorous statements with a reference to an exact assumption. Moreover, the above assumption contains more.\nc]

\begin{remark}
The lower bound $\rho_2\ge 1+\tau$ in equation \eqref{assm2} is to ensure that the sample covariance matrix $(X^{(2)})^\top X^{(2)}$ is non-singular with overwhelming probability. Moreover, we have allowed that $\rho_1=0$, because we want to state a general result that also covers the special case with no source task.
\end{remark}

\begin{remark}\label{rem_athmoment}
Whenever the estimates in equation \eqref{eq_support} hold, we say $n^{-1/2}Z^{(1)}$ and $n^{-1/2}Z^{(2)}$ satisfy the {\it{bounded support condition}} with $q$, or $n^{-1/2}Z^{(1)}$ and $n^{-1/2}Z^{(2)}$ have support $q$.
%\begin{definition}\label{defn_support}
%We say a (rescaled) random matrix $\cal M$ satisfies the {\it{bounded support condition}} with $q$, if
%\begin{equation}
%\max_{i,j}\vert \cal M_{ij}\vert \prec q. \label{eq_support}
%\end{equation}
%%Here $q\equiv q(n)$ is a deterministic parameter and usually satisfies $ n^{-{1}/{2}} \leq q \leq n^{- \phi} $ for some (small) constant $\phi>0$.
%Whenever (\ref{eq_support}) holds, we say that $M$ has support $q$.
%%Moreover, if the entries of $X$ satisfy (\ref{size_condition}), then $X$ trivially satisfies the bounded support condition with $q=N^{-\phi}$.
%\end{definition}
%\begin{example}
%\todo{Talk about the Gaussian case and Bernoulli case.}
By Example \ref{example stoc}, if the entries of $Z^{(1)}$ have finite moments up to any order, %as in equation \eqref{assmAhigh},
then $n^{-1/2}Z^{(1)}$ has bounded support $n^{-1/2}$. More generally, if the entries of $Z^{(1)}$ have finite $\varphi$-th moment as in \eqref{assmAhigh}, then using Markov's inequality and a simple union bound we get that
\be\label{Ptrunc}
\P\left(n^{-1/2}\max_{i,\mu}|(Z^{(1)})_{\mu i}|\ge (\log n) n^{-\frac{\varphi-4}{2\varphi}}\right)=\OO((\log n)^{-\varphi}).\ee
In other words, $n^{-1/2}Z^{(1)}$ has bounded support $q=n^{-\frac{\varphi-4}{2\varphi}}$ on a event with probability $1-\oo(1)$. This leads to the exponent $\frac{\varphi-4}{2\varphi}$ in Fact \ref{lem_minv}. %This fact will be used in the proof of Corollary \ref{main_cor} below.
The above discussion shows that the bounded support condition provides great flexibility in dealing with finite moment conditions.
%is essentially introduced to deal with the moment conditions; see Corollary \ref{main_cor} below for more details.
%\end{example}
\end{remark}

@@
